---
title: "ECON185_Assignment2_v1"
format: html
editor: visual
---

## Packages and Data

```{r}
#install.packages("readxl") 

library(tidyverse)
library(dplyr)
library(readxl)
library(ggplot2)
library(lubridate)
```

```{r}
setwd("C:/Users/ariel/Downloads/UCSD/WI 26/ECON 185")

df <- read_excel("RIPA Stop Data _ Santa Clara 2019.xlsx")

colnames(df) <- tolower(colnames(df))
```

## 1. Context and Summary Statistics

Task: Provide a table showing the total number of stops, the percentage of stops by race, and the reasons for the stops (not all reasons, pick a few that interest you).

```{r}
# changing race code to strings
#converting RAE_FULL codes to readable strings
df <- df %>%
  mutate(race_string = recode(rae_full,
    `1` = "Asian",
    `2` = "Black",
    `3` = "Hispanic/Latino",
    `4` = "MiddleEastern/SouthAsian",
    `5` = "NativeAmerican",
    `6` = "PacificIslander",
    `7` = "White",
    `8` = "Multiracial"))

# total number of stop
dim(df)[1]
```

```{r}
df %>% 
  group_by(race_string) %>% 
  summarise(tot_stop = n(),
            pct_stop = round(n()/dim(df)[1]*100,1),
            reason_traf = sum(reason_for_stop==1)/tot_stop,
            reason_sus = sum(reason_for_stop==2)/tot_stop,
            reason_srch = sum(reason_for_stop==6)/tot_stop)
```

```{r}
df %>% 
  filter(reason_for_stop == 1) %>% 
  group_by(rae_full) %>% 
  summarise(tot_stop = n(),
            pct_stop = round(n()*100/(sum(df$reason_for_stop==1)), 2))
```

● Task: Create a visualization showing the Search Rate (Searches / Total Stops) across different racial groups. (There might be a few definitions of “search” - with warrant, without warrant, with consent, without consent etc. - be specific, or create an aggregate “any search” variable from these).\

```{r}
# aggregate var: search of person/property
df = df %>% 
  mutate(any_search = ifelse(
    ads_search_person==1 | ads_search_property==1, TRUE, FALSE))

# df %>% summarise(any_search_count = sum(any_search))
```

```{r}
df %>% 
  group_by(race_string) %>% 
  summarise(search_rate = 100*sum(any_search==TRUE)/n()) %>% 
  ggplot(aes(x = race_string, y = search_rate)) +
  geom_col()
```

```{r}
# gemini improved plot

df %>% 
  group_by(race_string) %>% 
  summarise(search_rate = 100 * sum(any_search == TRUE) / n()) %>% 
  ggplot(aes(x = reorder(race_string, -search_rate), y = search_rate)) +
  geom_col(fill = "darkblue", width = 0.7) +
  # Adding the labels
  geom_text(
    aes(label = sprintf("%.1f%%", search_rate)), # Rounds to 1 decimal and adds % sign
    vjust = -0.5,                                # Nudges text slightly above the bar
    size = 3,                                    # Adjust font size
    fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + # Adds room at the top for labels
  theme_minimal(base_size = 14) +
  labs(
    title = "Search Rate by Racial Group",
    x = NULL,                                    # Removing x-axis title as labels are self-explanatory
    y = "Search Rate (%)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 9),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", margin = margin(b = 10))
  )
```

```{r}
df %>% 
  #group_by(race_string) %>% 
  summarise(tot_cons_search = sum(reason_for_stop==6))
```

```{r}
df %>% 
  group_by(race_string) %>% 
  summarise(
    tot_search = sum(any_search==TRUE),
    pct_byrace_search = tot_search/n()*100)
```

```{r}
df %>% 
  group_by(race_string) %>% 
  summarise(
    tot_search = sum(any_search==TRUE),
    pct_all_search = tot_search/13832*100)
```

● Question: Does a simple disparity in search rates prove discrimination? Why or why not

### The Veil of Darkness (VOD) Test

The VOD test assumes that police officers cannot easily identify a driver's race at night before making a stop.If the share of minority drivers stopped is higher during daylight than after dark, it would be consistent with racial profiling.

● Task: Estimate a regression where the dependent variable is an indicator for a Black/Hispanic driver being stopped and present your results in a table.

○ Key Variable: An indicator for "Daylight" (is it light or dark at the time of the stop?).

○ Constraint: You must control for the time of day (e.g., in 15-minute "bins") to ensure you aren't just picking up on different driving patterns at night. Focus on the margin of when the sun might be setting.

● Question: Does the probability of stopping a minority driver increase when the sun is up?

```{r}
# making new var for race=black/brown
df = df %>% 
  mutate(
    blackbrown = ifelse(
      rae_full==2|rae_full==3, TRUE, FALSE))
```

```{r}
# controlling for time with 15 minute bins
# gemini code

df <- df %>%
  mutate(
    # 1. Pad the number with leading zeros so it is always 4 digits (e.g., 22 becomes "0022")
    time_str = sprintf("%04d", time_of_stop),
    
    # 2. Convert that string into a proper time object
    # We use a dummy date because POSIXct objects always require a date component
    time_obj = as.POSIXct(time_str, format = "%H%M"),
    
    # 3. Create 15-minute bins 
    # This now works because time_obj is a POSIXct class
    time_bin = cut(time_obj, breaks = "15 min")
  )

# Verify the bins
summary(df$time_bin)
```

```{r}
# making new one hot encoded var for sun up/down
# darkness after 7pm
df <- df %>% 
  mutate(
    sundown = ifelse(time_of_stop>1900, TRUE, FALSE),
    dark = ifelse(time_of_stop>1900|time_of_stop<500, TRUE, FALSE))
```

```{r}
mod = glm(data = df, blackbrown ~ dark + factor(time_bin), family = binomial) 
summary(mod)
```

```{r}
df %>% 
  group_by(dark, race_string) %>% 
  summarise(count = n())
```

```{r}
df %>% 
  group_by(blackbrown) %>% 
  summarise(count = n())
```

```{r}
# Log-odds (intercept)
log_odds <- mod[["coefficients"]][["(Intercept)"]]

# Convert to probability
p <- exp(log_odds) / (1 + exp(log_odds))
p
```

```{r}
# Log-odds (darkTRUE)
log_odds <- mod[["coefficients"]][["(Intercept)"]] + mod[["coefficients"]][["darkTRUE"]]

# Convert to probability
p <- exp(log_odds) / (1 + exp(log_odds))
p
```

### 3. Becker's Outcome Test: Hit Rates

```{r}
# black/brown vs white var, df of only black/brown/white
df_bbw = df %>% 
  filter(rae_full %in% c(7,2,3)) %>%  
  mutate(blackbrown_white = ifelse(
    rae_full==2|rae_full==3, TRUE, FALSE))

#removing NA row under ced_none_contraband
df_bbw = subset(df_bbw, complete.cases(ced_none_contraband))
```

```{r}
df_bbw %>% 
  group_by(blackbrown_white) %>% 
  summarise(hit_search_rate = sum(ced_none_contraband==0)/sum(any_search==TRUE))
```

```{r}
# making contraband var more intuitive to read
df_bbw = df_bbw %>% 
  group_by(blackbrown_white) %>% 
  mutate(hit = ifelse(ced_none_contraband==0, TRUE, FALSE))
```

```{r}
# groupby blackbrown_white, tot_search, tot_hit, hit_rate
df_bbw %>% 
  group_by(blackbrown_white) %>% 
  summarise(
            #tot_stop = n(),
            tot_search = sum(any_search==1),
            search_rate = tot_search/n(),
            tot_hit = sum(hit==TRUE),
            hit_rate = tot_hit/tot_search)
```

```{r}
# subset df to only searched individuals
df_bbw_searched = df_bbw %>% 
  filter(any_search==1)
```

```{r}
mod2 = glm(data = df_bbw_searched, hit ~ blackbrown_white, family = binomial)
summary(mod2)
```

```{r}
# Log-odds (intercept)
log_odds <- mod2[["coefficients"]][["(Intercept)"]]

# Convert to probability
p <- exp(log_odds) / (1 + exp(log_odds))
p
```

```{r}
# Log-odds (blackbrown_whiteTRUE)
log_odds <- mod2[["coefficients"]][["(Intercept)"]] + mod2[["coefficients"]][["blackbrown_whiteTRUE"]]

# Convert to probability
p <- exp(log_odds) / (1 + exp(log_odds))
p
```

\
